---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The following code reads in the data into our data frame 'activity'.
```{r, results='hide'}
activity <- read.csv("activity.csv")
```

Here I also created a subset of the data with the NA's removed:
```{r, results = 'hide'}
activity_noNA <- activity[!is.na(activity$steps),]
```


## What is mean total number of steps taken per day?
The following code creates a table (called 'date_sum') of the sum of the number of steps per day.
```{r}
library(plyr)
date_sum <- ddply(activity_noNA, .(date), summarize, sum_steps=sum(steps))
date_sum
```
This code creates the histogram:
```{r First Histogram}
hist(date_sum$sum_steps,breaks=50,freq=TRUE,xlab = "Steps per day", ylab = "Frequency", main = "Histogram of Steps Taken Per Day")
```

To calculate the mean of the number of steps taken each day:
```{r}
ddply(activity_noNA, .(date), summarize, mean_steps=mean(steps))

```
and the median.
```{r}
ddply(activity_noNA, .(date), summarize, median_steps=median(steps))
```


## What is the average daily activity pattern?
For this section, I also used the data set 'activity_noNA', i.e. the data set with NA's removed.  The following code creates a plot of the average number of steps taken during each 5 minute time interval (0,5,10,...2355) for all days.
```{r First Time Plot}
interval_mn<-ddply(activity_noNA, .(interval), summarize,      mean_steps = mean(steps))
plot(interval_mn$interval, interval_mn$mean_steps,type="l",  main = "Average Number of Steps per Time Interval", xlab = "Time Interval",  ylab = "Average number of steps")
```

The time interval with the max average number of steps taken over all the days can be found with this code:
```{r}
interval_mn[interval_mn$mean_steps==max(interval_mn$mean_steps),]
```
and it is the 5 minutes following 8:35 AM.


## Imputting missing values
The total number of NA's in the data set is 2304, found using the following code:
```{r}
sum(is.na(activity$steps))
```

I decided to replace the NA's with the average steps over all days for each interval. The following code creates a new data set ('new_activity') and replaces each NA value.
```{r}
new_activity <- activity
for (i in 1:length(new_activity$steps)){
     if (is.na(new_activity[i,1])){
          new_activity[i,1] <- interval_mn[interval_mn$interval==new_activity[i,3],2]
     }
}
```
We can check that it worked:
```{r}
head(activity)
head(new_activity)
```

Here is the histogram of our data with the NA's replaced:
```{r Second Histogram}
new_date_sum <- ddply(new_activity, .(date), summarize, sum_steps=sum(steps))
hist(new_date_sum$sum_steps,breaks=50,freq=TRUE,xlab = "Steps per day", ylab = "Frequency", main = "Histogram of Steps Taken Per Day (NA's replaced)")
```

Replacing the NA's only changed the middle 3 ranges of the histogram.
Here are the new mean data for each day:
```{r}
ddply(new_activity, .(date), summarize, mean_steps=mean(steps))

```
This adds in the dates we lost in removing the NA's from the previous data.

This is the new median data:
```{r}
ddply(new_activity, .(date), summarize, median_steps=median(steps))
```
Since here we're just adding in data for the missing dates, the median should be the median of the interval means we calculated.

## Are there differences in activity patterns between weekdays and weekends?

The following code changes our existing dates to POSIXlt format, finds the weekdays corresponding to the dates, and then creates a new column in 'new_activity' called 'class' with classifications of weekend and weekday.
```{r}
#Changing to date format
dates <- strptime(new_activity$date,format="%Y-%m-%d")
#Labeling days
class <- weekdays(dates)
#Creating new column
new_activity$class<-class
#Renaming Weekends and Weekdays
new_activity[(new_activity$class=="Sunday"|new_activity$class=="Saturday"),4]<-"Weekend"
new_activity[!(new_activity$class=="Weekend"),4]<-"Weekday"
#Turning column into factors
new_activity$class<-factor(new_activity$class)
```

Here we average over all weekends and weekdays for each interval and the use the lattice system to plot:
```{r Second Time Plot}
class_mn<-ddply(new_activity, .(interval,class), summarize, mean_steps = mean(steps))
library(lattice)
xyplot(mean_steps ~ interval | class, data=class_mn, type = "l",layout = c(1,2), ylab= "Average Steps Taken", xlab = "Interval")
```